---
title: "Exploratory Analysis"
author: "Elling Payne"
date: "2025-04-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE);
knitr::opts_chunk$set(fig.align = "center");
knitr::opts_chunk$set(seed = 112358);
library(tidyverse);
# library(e1071);
```

## Where did the data come from and what does it represent? What motivations are there for exploring it?

Responses to the IPUMS Health Surveys: National Health Interview Survey

## How much data is there? What are the valid entries for each feature?

```{r}
ipums <- read.csv("input/nhis_2022.csv");

# remove some variables that will not be useful or interesting for subsetting or modeling
clean_ipums <- data.frame(ipums);
total_rows_dirty <- nrow(ipums);
survey_info_cols <- names(ipums)[1:12];
demographic_cols <- names(ipums)[c(13:18, 22)];
disease_indicator_cols <- names(ipums)[23:27];
health_cols <- names(ipums)[c(19:21, 28:ncol(ipums))];
print(survey_info_cols);
print(demographic_cols);
print(disease_indicator_cols);
print(health_cols);
```

```{r}
# assumes data has already been cast to numeric where applicable
# converts codes representing missing data to NA and converts HRSLEEP code representing less than
# one hour to 0 so that the new HRSLEEP is 0-24.
# For flags indicating a sample adult or sample child, they are converted to NA if not the sample
# or the sample but data is not available
codes_to_na <- function(data) {
  # fields that are yearly or monthly consumption times and
  # >996 is a missing value (995 = 995 or more)
  more_freq_consume <- c("FRUTNO", "VEGENO", "BEANNO", "SODAPNO", "FRIESPNO",
                         "SPORDRMNO", "FRTDRINKMNO", "COFETEAMNO", "POTATONO",
                         "PIZZANO", "SALSAMNO", "TOMSAUCEMNO");
  # for diseases besides diabetes, 1 is yes ever had, 2 is no never had
  # for diabetes, 3 is valid and means borderline had
  disease_not_diab <- c("CANCEREV", "CHEARTDIEV", "HEARTATTEV", "STROKEV");
  # fields that are yearly or monthly consumption but have a max value of 95
  less_freq_consume <- c("JUICEMNO", "SALADSNO");
  clean_data <- data.frame(data);
  clean_data[, more_freq_consume] <- clean_data[, more_freq_consume] %>%
    lapply(function(x) {ifelse(x > 995 | x < 0 | is.na(x), NA, x)});
  clean_data[, less_freq_consume] <- clean_data[, less_freq_consume] %>%
    lapply(function(x) {ifelse(x > 95 | x < 0 | is.na(x), NA, x)});
  clean_data[, disease_not_diab] <- clean_data[, disease_not_diab] %>%
    lapply(function(x) {ifelse(x == 1 | x == 2 | is.na(x), x, NA)});
  clean_data <- clean_data %>%
    mutate(DIABETICEV = ifelse(is.nan(DIABETICEV) | 
                                 (DIABETICEV != 1 & DIABETICEV != 2 & DIABETICEV != 3),
                               NA, DIABETICEV)) %>%
    mutate(HOURSWRK = ifelse(HOURSWRK > 96 | HOURSWRK < 0 | is.na(HOURSWRK), NA, HOURSWRK)) %>%
    mutate(HRSLEEP = ifelse(HRSLEEP > 25 | HRSLEEP < 1 | is.na(HRSLEEP), NA, 
                            ifelse(HRSLEEP == 25, 0, HRSLEEP))) %>%
    mutate(BMICALC = ifelse(BMICALC == 996 | is.na(BMICALC), NA, BMICALC)) %>%
    mutate(AGE = ifelse(AGE > 99 | is.na(AGE), NA, AGE)) %>%
    mutate(SEX = ifelse((SEX != 1 & SEX != 2) | is.na(SEX), NA, SEX)) %>%
    mutate(ASTATFLG = ifelse(ASTATFLG != 1 | is.na(ASTATFLG), 0, 1)) %>%
    mutate(CSTATFLG = ifelse(CSTATFLG != 1 | is.na(CSTATFLG), 0, 1)) %>%
    mutate(ALCDAYSYR = ifelse(ALCDAYSYR > 365 | ALCDAYSYR < 0 | is.na(ALCDAYSYR),
                              NA, ALCDAYSYR)) %>%
    mutate(CIGDAYMO = ifelse(is.na(CIGDAYMO) | CIGDAYMO < 0 | CIGDAYMO > 30,
                             NA, CIGDAYMO)) %>%
    mutate(MOD10DMIN = ifelse(is.na(MOD10DMIN) | MOD10DMIN < 0 | MOD10DMIN > 719,
                              NA, MOD10DMIN)) %>%
    mutate(VIG10DMIN = ifelse(is.na(VIG10DMIN) | VIG10DMIN < 0 | VIG10DMIN > 719,
                              NA, VIG10DMIN));
  
  # other demographic variables are likely not valid to use in SVM without vecotrization
  # since they break interval assumptions making distance comparison difficult.
  # won't bother cleaning them since I will not be using them
  # weight might be usable, but is closely related to BMI so I will leave it out for now
  # some of the binary variables like health insurance status might be usable
  # but may not be as interesting for a SVM
  return(clean_data);
}

# first remove some columns I am not interested in
clean_ipums <- clean_ipums %>%
  select(-c("SERIAL", "MARSTCUR", "STRATA", "PSU", "NHISHID", "REGION", "PERNUM", "NHISPID",
            "HHX", "SAMPWEIGHT", "EDUC", "HEIGHT", "WEIGHT", "HINOTCOVE", 
            "ALCANYNO", "CVDSHT", "POVERTY")) %>%
  lapply(as.numeric); # those categories left are numerically encoded

clean_ipums <- codes_to_na(clean_ipums);
```

```{r}
names(clean_ipums);
```
```{r}
numeric_cols = c("YEAR", "AGE", "HOURSWRK", "BMICALC", "ALCDAYSYR", "CIGDAYMO",
                 "MOD10DMIN", "VIG10DMIN", "FRUTNO", "VEGENO", "JUICEMNO", "SALADSNO",
                 "BEANNO", "SALSAMNO", "TOMSAUCEMNO", "SODAPNO", "FRIESPNO",
                 "SPORDRMNO", "FRTDRINKMNO", "COFETEAMNO", "POTATONO", "PIZZANO",
                 "HRSLEEP", "HOURSLP");
factor_cols = c("ASTATFLG", "CSTATFLG", "SEX", "CANCEREV",
                "CHEARTDIEV", "DIABETICEV", "HEARTATTEV", "STROKEV");
clean_ipums[, factor_cols] <- clean_ipums[, factor_cols] %>% lapply(as.factor);
```



## How much data is missing? Is there anything of note about which data is missing?

```{r}
print(paste0("Total Rows in Dataset: ", nrow(clean_ipums)));
clean_ipums %>% lapply(function(x) {sum(is.na(x))});
```
It appears that all of the data here is either from a sample adult or sample child, so we don't have to worry about nonsample participants. Some of the other variables are missing up to around around a quarter of the records
```{r}
# even though 3 was a valid class for DIABETICEV, none are 3 so it can be treated as binary (yay)
table(clean_ipums$DIABETICEV);
```


```{r}
# lets subset to sample adult men with info. Adults because a lot of data is only collected for
# sample adults and men just to reduce the sample size
clean_ipums <- clean_ipums[!is.na(clean_ipums$ASTATFLG) & clean_ipums$ASTATFLG == 1,];
clean_ipums <- clean_ipums[!is.na(clean_ipums$SEX) & clean_ipums$SEX == 1,];
clean_ipums <- clean_ipums %>% select(-c("CSTATFLG", "SEX"))
nrow(clean_ipums);
```
```{r}
# now we have good data for most of the interesting features
clean_ipums %>% lapply(function(x) {sum(is.na(x))});
```
```{r}
# CIGDAYMO is missing for most of the subset, so lets get rid of that right now
clean_ipums <- clean_ipums %>% select(-CIGDAYMO);
```



## What is the distribution of each feature and target? What imbalances or other concerns are there?

```{r}
# possible responses
responses <- c("CANCEREV", "CHEARTDIEV", "DIABETICEV", "HEARTATTEV", "STROKEV");
features <- c("AGE", "HOURSWRK", "BMICALC", "ALCDAYSYR", "MOD10DMIN",
              "VIG10DMIN", "FRUTNO", "VEGENO", "JUICEMNO", "SALADSNO", "BEANNO",
              "SALSAMNO", "TOMSAUCEMNO", "SODAPNO", "FRIESPNO", "SPORDRMNO",
              "FRTDRINKMNO", "COFETEAMNO", "POTATONO", "PIZZANO", "HRSLEEP");
```
```{r}
for (response in responses) {
  plt <- ggplot(data=clean_ipums, mapping=aes(x=.data[[response]])) +
    geom_bar(fill="skyblue", color="darkgrey");
  print(plt);
  ggsave(paste0("output/eda/plots/", response, "_class_frequencies_bar.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```
```{r}
for (feature in features) {
  plt <- ggplot(data = clean_ipums, mapping = aes(x=.data[[feature]])) +
    geom_histogram(bins=30, fill="skyblue", color="darkgrey");
  print(plt);
  ggsave(paste0("output/eda/plots/", feature, "_histogram.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```




## plots of target class vs features

```{r}
# to get an idea of what features and responses might be most interesting,
# lets remove na rows then plot the features vs responses
na_removed <- clean_ipums %>% na.omit();
for (response in responses) {
  for (feature in features) {
    plt <- ggplot(data = na_removed, mapping = aes(x=.data[[response]], y=.data[[feature]])) +
      geom_boxplot(position = "dodge", fill="skyblue", color="darkgrey") +
      labs(title=paste0(response, " ~ ", feature));
    print(plt);
    ggsave(paste0("output/eda/plots/", response, "_vs_", feature, "_boxplot.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
  }
}
```

```{r}
# DIABETICEV looked interesting, lets do the plots again for that without outliers
# to get an idea of what features and responses might be most interesting,
# lets remove na rows then plot the features vs responses
for (feature in features) {
  inter_quartile_range <- IQR(na_removed[[feature]])
  limits <- c(quantile(na_removed[[feature]], 0.25) - 1.5*inter_quartile_range,
              quantile(na_removed[[feature]], 0.75) + 1.5*inter_quartile_range)
  plt <- ggplot(data = na_removed, mapping = aes(x=DIABETICEV, y=.data[[feature]])) +
    geom_boxplot(position = "dodge", fill="skyblue", color="darkgrey",
                 outlier.shape = NA) +
    coord_cartesian(ylim = limits) +
    labs(title=paste0( "DIABETICEV ~ ", feature));
  print(plt);
  ggsave(paste0("output/eda/plots/", response, "_vs_", feature, "_boxplot_no_outliers.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```


## pairwise scatter plots of features vs the reponse in color

```{r}
col_is_numeric <- sapply(na_removed, is.numeric);
normalized_data <- data.frame(na_removed);
normalized_data[, col_is_numeric] <- na_removed[, col_is_numeric] %>%
  lapply(function(x) { return((x - mean(x))/sd(x)) });

for (response in responses) {
  for (pair in combn(features, 2, simplify = FALSE)) {
    plt <- ggplot(data = normalized_data, mapping = aes(x=.data[[pair[1]]],
                                                    y=.data[[pair[2]]],
                                                    color=.data[[response]])) +
      geom_point() + 
      labs(title = paste0(response, " ~ ", pair[1], " + ", pair[2], " (Normalized)"),
           x = paste0("Normalized ", pair[1]), y = paste0("Normalized ", pair[2]),
           color = response);
    scale_labels <- c("No", "Yes");
    scale_values <- c("1"="skyblue", "2"="coral");
    if (response == "DIABETICEV") {
      scale_labels <- c(scale_labels, "Borderline");
      scale_values <- c(scale_values, "3"="limegreen")
    }
    plt <- plt + scale_color_manual(values = scale_values,
                                    labels = scale_labels);
    print(plt);
    ggsave(paste0("output/eda/plots/", response, "_vs_", pair[1], "_and_", pair[2],
                  "_scatter.png"), plot = plt, width = 8, height = 8, dpi = "print");
  }
}
```


## Selection of final features and responses to explore

```{r}
# select DIABETICEV as the response
# (Even though it has three valid categories, only two are present so its still a single SVM)
# And select some of the more interesting features based on Box plots and scatter plots
diabetes_ipums <- clean_ipums %>% select(c("DIABETICEV", "AGE", "BMICALC", "ALCDAYSYR",
                                           "HOURSWRK", "SODAPNO", "FRIESPNO")) %>%
  na.omit();
write.csv(diabetes_ipums, file="output/preprocessing/data/selected_response_and_features.csv",
          row.names = FALSE);
save(diabetes_ipums, file="output/preprocessing/data/selected_response_and_features.RData");
```

