---
title: "Exploratory Analysis"
author: "Elling Payne"
date: "2025-05-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE);
knitr::opts_chunk$set(root.dir = dirname(rstudioapi::getActiveDocumentContext()$path));
knitr::opts_chunk$set(fig.align = "center");
knitr::opts_chunk$set(seed = 112358);
library(tidyverse);
```

## Where did the data come from and what does it represent? What motivations are there for exploring it?

Responses to the IPUMS Health Surveys: National Health Interview Survey

## How much data is there? What are the valid entries for each feature?

```{r}
ipums <- read.csv("input/nhis_2022.csv");

# remove some variables that will not be useful or interesting for subsetting or modeling
clean_ipums <- data.frame(ipums);
total_rows_dirty <- nrow(ipums);
survey_info_cols <- names(ipums)[1:12];
demographic_cols <- names(ipums)[c(13:18, 22)];
disease_indicator_cols <- names(ipums)[23:27];
health_cols <- names(ipums)[c(19:21, 28:ncol(ipums))];
print(survey_info_cols);
print(demographic_cols);
print(disease_indicator_cols);
print(health_cols);
```

```{r}
# assumes data has already been cast to numeric where applicable
# converts codes representing missing data to NA and converts HRSLEEP code representing less than
# one hour to 0 so that the new HRSLEEP is 0-24.
# For flags indicating a sample adult or sample child, they are converted to NA if not the sample
# or the sample but data is not available
codes_to_na <- function(data) {
  # fields that are yearly or monthly consumption times and
  # >996 is a missing value (995 = 995 or more)
  more_freq_consume <- c("FRUTNO", "VEGENO", "BEANNO", "SODAPNO", "FRIESPNO",
                         "SPORDRMNO", "FRTDRINKMNO", "COFETEAMNO", "POTATONO",
                         "PIZZANO", "SALSAMNO", "TOMSAUCEMNO");
  # for diseases besides diabetes, 1 is yes ever had, 2 is no never had
  # for diabetes, 3 is valid and means borderline had
  disease_not_diab <- c("CANCEREV", "CHEARTDIEV", "HEARTATTEV", "STROKEV");
  # fields that are yearly or monthly consumption but have a max value of 95
  less_freq_consume <- c("JUICEMNO", "SALADSNO");
  clean_data <- data.frame(data);
  clean_data[, more_freq_consume] <- clean_data[, more_freq_consume] %>%
    lapply(function(x) {ifelse(x > 995 | x < 0 | is.na(x), NA, x)});
  clean_data[, less_freq_consume] <- clean_data[, less_freq_consume] %>%
    lapply(function(x) {ifelse(x > 95 | x < 0 | is.na(x), NA, x)});
  clean_data[, disease_not_diab] <- clean_data[, disease_not_diab] %>%
    lapply(function(x) {ifelse(x == 1 | x == 2 | is.na(x), x, NA)});
  clean_data <- clean_data %>%
    mutate(DIABETICEV = ifelse(is.nan(DIABETICEV) | 
                                 (DIABETICEV != 1 & DIABETICEV != 2 & DIABETICEV != 3),
                               NA, DIABETICEV)) %>%
    mutate(HOURSWRK = ifelse(HOURSWRK > 96 | HOURSWRK < 0 | is.na(HOURSWRK), NA, HOURSWRK)) %>%
    mutate(HRSLEEP = ifelse(HRSLEEP > 25 | HRSLEEP < 1 | is.na(HRSLEEP), NA, 
                            ifelse(HRSLEEP == 25, 0, HRSLEEP))) %>%
    mutate(BMICALC = ifelse(BMICALC == 996 | is.na(BMICALC), NA, BMICALC)) %>%
    mutate(AGE = ifelse(AGE > 99 | is.na(AGE), NA, AGE)) %>%
    mutate(SEX = ifelse((SEX != 1 & SEX != 2) | is.na(SEX), NA, SEX)) %>%
    mutate(ASTATFLG = ifelse(ASTATFLG != 1 | is.na(ASTATFLG), 0, 1)) %>%
    mutate(CSTATFLG = ifelse(CSTATFLG != 1 | is.na(CSTATFLG), 0, 1)) %>%
    mutate(ALCDAYSYR = ifelse(ALCDAYSYR > 365 | ALCDAYSYR < 0 | is.na(ALCDAYSYR),
                              NA, ALCDAYSYR)) %>%
    mutate(CIGDAYMO = ifelse(is.na(CIGDAYMO) | CIGDAYMO < 0 | CIGDAYMO > 30,
                             NA, CIGDAYMO)) %>%
    mutate(MOD10DMIN = ifelse(is.na(MOD10DMIN) | MOD10DMIN < 0 | MOD10DMIN > 719,
                              NA, MOD10DMIN)) %>%
    mutate(VIG10DMIN = ifelse(is.na(VIG10DMIN) | VIG10DMIN < 0 | VIG10DMIN > 719,
                              NA, VIG10DMIN));
  
  # other demographic variables are likely not valid to use in SVM without vecotrization
  # since they break interval assumptions making distance comparison difficult.
  # won't bother cleaning them since I will not be using them
  # weight might be usable, but is closely related to BMI so I will leave it out for now
  # some of the binary variables like health insurance status might be usable
  # but may not be as interesting for a SVM
  return(clean_data);
}

# first remove some columns I am not interested in
clean_ipums <- clean_ipums %>%
  select(-c("SERIAL", "MARSTCUR", "STRATA", "PSU", "NHISHID", "REGION", "PERNUM", "NHISPID",
            "HHX", "SAMPWEIGHT", "EDUC", "HEIGHT", "WEIGHT", "HINOTCOVE", 
            "ALCANYNO", "CVDSHT", "POVERTY")) %>%
  lapply(as.numeric); # those categories left are numerically encoded

clean_ipums <- codes_to_na(clean_ipums);
```

```{r}
names(clean_ipums);
```
```{r}
numeric_cols = c("YEAR", "AGE", "HOURSWRK", "BMICALC", "ALCDAYSYR", "CIGDAYMO",
                 "MOD10DMIN", "VIG10DMIN", "FRUTNO", "VEGENO", "JUICEMNO", "SALADSNO",
                 "BEANNO", "SALSAMNO", "TOMSAUCEMNO", "SODAPNO", "FRIESPNO",
                 "SPORDRMNO", "FRTDRINKMNO", "COFETEAMNO", "POTATONO", "PIZZANO",
                 "HRSLEEP", "HOURSLP");
factor_cols = c("ASTATFLG", "CSTATFLG", "SEX", "CANCEREV",
                "CHEARTDIEV", "DIABETICEV", "HEARTATTEV", "STROKEV");
clean_ipums[, factor_cols] <- clean_ipums[, factor_cols] %>% lapply(as.factor);
```



## How much data is missing? Is there anything of note about which data is missing?

```{r}
print(paste0("Total Rows in Dataset: ", nrow(clean_ipums)));
clean_ipums %>% lapply(function(x) {sum(is.na(x))});
```
It appears that all of the data here is either from a sample adult or sample child, so we don't have to worry about nonsample participants. Some of the other variables are missing up to around around a quarter of the records
```{r}
# even though 3 was a valid class for DIABETICEV, none are 3 so it can be treated as binary (yay)
table(clean_ipums$DIABETICEV);
```


```{r}
# lets subset to sample adult men with info. Adults because a lot of data is only collected for
# sample adults and men just to reduce the sample size
clean_ipums <- clean_ipums[!is.na(clean_ipums$ASTATFLG) & clean_ipums$ASTATFLG == 1,];
clean_ipums <- clean_ipums[!is.na(clean_ipums$SEX) & clean_ipums$SEX == 1,];
clean_ipums <- clean_ipums %>% select(-c("CSTATFLG", "SEX"))
nrow(clean_ipums);
```
```{r}
# now we have good data for most of the interesting features
clean_ipums %>% lapply(function(x) {sum(is.na(x))});
```
```{r}
# CIGDAYMO is missing for most of the subset, so lets get rid of that right now
clean_ipums <- clean_ipums %>% select(-CIGDAYMO);
```



## What is the distribution of each feature and target? What imbalances or other concerns are there?

```{r}
# possible responses
responses <- c("CANCEREV", "CHEARTDIEV", "DIABETICEV", "HEARTATTEV", "STROKEV");
features <- c("AGE", "HOURSWRK", "BMICALC", "ALCDAYSYR", "MOD10DMIN",
              "VIG10DMIN", "FRUTNO", "VEGENO", "JUICEMNO", "SALADSNO", "BEANNO",
              "SALSAMNO", "TOMSAUCEMNO", "SODAPNO", "FRIESPNO", "SPORDRMNO",
              "FRTDRINKMNO", "COFETEAMNO", "POTATONO", "PIZZANO", "HRSLEEP");
```
```{r}
for (response in responses) {
  plt <- ggplot(data=clean_ipums, mapping=aes(x=.data[[response]])) +
    geom_bar(fill="skyblue", color="darkgrey");
  print(plt);
  ggsave(paste0("output/eda/plots/", response, "_class_frequencies_bar.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```
```{r}
for (feature in features) {
  plt <- ggplot(data = clean_ipums, mapping = aes(x=.data[[feature]])) +
    geom_histogram(bins=30, fill="skyblue", color="darkgrey");
  print(plt);
  ggsave(paste0("output/eda/plots/", feature, "_histogram.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```




## plots of target class vs features

```{r}
# to get an idea of what features and responses might be most interesting,
# lets remove na rows then plot the features vs responses
na_removed <- clean_ipums %>% na.omit();
for (response in responses) {
  for (feature in features) {
    plt <- ggplot(data = na_removed, mapping = aes(x=.data[[response]], y=.data[[feature]])) +
      geom_boxplot(position = "dodge", fill="skyblue", color="darkgrey") +
      labs(title=paste0(response, " ~ ", feature));
    print(plt);
    ggsave(paste0("output/eda/plots/", response, "_vs_", feature, "_boxplot.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
  }
}
```

```{r}
# DIABETICEV looked interesting, lets do the plots again for that without outliers
# to get an idea of what features and responses might be most interesting,
# lets remove na rows then plot the features vs responses
for (feature in features) {
  inter_quartile_range <- IQR(na_removed[[feature]])
  limits <- c(quantile(na_removed[[feature]], 0.25) - 1.5*inter_quartile_range,
              quantile(na_removed[[feature]], 0.75) + 1.5*inter_quartile_range)
  plt <- ggplot(data = na_removed, mapping = aes(x=DIABETICEV, y=.data[[feature]])) +
    geom_boxplot(position = "dodge", fill="skyblue", color="darkgrey",
                 outlier.shape = NA) +
    coord_cartesian(ylim = limits) +
    labs(title=paste0( "DIABETICEV ~ ", feature));
  print(plt);
  ggsave(paste0("output/eda/plots/", response, "_vs_", feature, "_boxplot_no_outliers.png"),
           plot = plt, width = 8, height = 8, dpi = "print");
}
```


## pairwise scatter plots of features vs the reponse in color

I briefly looked over scatter plots for all of the interesting features that were easily translatable into interval variables. As the file became to large to upload to github with all of the plots, a small sample has been included here. In general, none of the plots show a strong obvious grouping with reasonable separation. This does not bode well for the use of a support vector machine unless some transformation could be identified which produces separation.

```{r, eval=FALSE}
col_is_numeric <- sapply(na_removed, is.numeric);
normalized_data <- data.frame(na_removed);
normalized_data[, col_is_numeric] <- na_removed[, col_is_numeric] %>%
  lapply(function(x) { return((x - mean(x))/sd(x)) });

for (response in responses) {
  for (pair in combn(features, 2, simplify = FALSE)) {
    plt <- ggplot(data = normalized_data, mapping = aes(x=.data[[pair[1]]],
                                                    y=.data[[pair[2]]],
                                                    color=.data[[response]])) +
      geom_point() + 
      labs(title = paste0(response, " ~ ", pair[1], " + ", pair[2], " (Normalized)"),
           x = paste0("Normalized ", pair[1]), y = paste0("Normalized ", pair[2]),
           color = response);
    scale_labels <- c("No", "Yes");
    scale_values <- c("1"="skyblue", "2"="coral");
    if (response == "DIABETICEV") {
      scale_labels <- c(scale_labels, "Borderline");
      scale_values <- c(scale_values, "3"="limegreen")
    }
    plt <- plt + scale_color_manual(values = scale_values,
                                    labels = scale_labels);
    print(plt);
    ggsave(paste0("output/eda/plots/", response, "_vs_", pair[1], "_and_", pair[2],
                  "_scatter.png"), plot = plt, width = 8, height = 8, dpi = "print");
  }
}
```
```{r}
# print 10 sample plots
col_is_numeric <- sapply(na_removed, is.numeric);
normalized_data <- data.frame(na_removed);
normalized_data[, col_is_numeric] <- na_removed[, col_is_numeric] %>%
  lapply(function(x) { return((x - mean(x))/sd(x)) });

fig_cap <- 10;
fig_cnt <- 0;
response = "DIABETICEV";
for (pair in combn(features, 2, simplify = FALSE)) {
  if (fig_cnt >= fig_cap) break;
  plt <- ggplot(data = normalized_data, mapping = aes(x=.data[[pair[1]]],
                                                  y=.data[[pair[2]]],
                                                  color=.data[[response]])) +
    geom_point() + 
    labs(title = paste0(response, " ~ ", pair[1], " + ", pair[2], " (Normalized)"),
         x = paste0("Normalized ", pair[1]), y = paste0("Normalized ", pair[2]),
         color = response);
  scale_labels <- c("No", "Yes");
  scale_values <- c("1"="skyblue", "2"="coral");
  if (response == "DIABETICEV") {
    scale_labels <- c(scale_labels, "Borderline");
    scale_values <- c(scale_values, "3"="limegreen")
  }
  plt <- plt + scale_color_manual(values = scale_values,
                                  labels = scale_labels);
  print(plt);
  ggsave(paste0("output/eda/plots/sample_", response, "_vs_", pair[1], "_and_", pair[2],
                "_scatter.png"), plot = plt, width = 8, height = 8, dpi = "print");
  fig_cnt <- fig_cnt + 1;
}
```



## Selection of final features and responses to explore

While none of the eda yielded greatly encouraging results, some of the boxplots for DIABETICEV appeared to show relationships. These included age, hours worked last week, bmi, days of alcohol consumption in the previous year, quantity of soda consumed, and quantity of fries consumed. There is some concern with the consumption features being interval variables. This is because some of the values are calculated from responses to the consumption question based on smaller time intervals than one year. However, the variables are still closer to interval variables in appearance and meaning than some of the other features such as poverty level. They might be considered close enough to attempt to build a SVM. Diabetes might also be a problematic choice given the significant class imbalance, which might make it hard for the model to learn the features of a positive diagnosis. A SVM will not be that sensitive to class imbalance if there is a good deal of separation in the classes, since only the support vectors matter. However, in the case of a lot of overlap and necessarily many support vectors, the contribution of the minority class will be diluted. Further, DIABETICEV appears to have three valid categories, but there are no examples of the "borderline" category. As such, it will be treated as a case of binary classification with the caveat that the model may not generalize well to those with borderline diabetes without further training. A consolidation of the borderline category into one of the others, or a multi-class model would be needed as well.

```{r}
# select DIABETICEV as the response
# (Even though it has three valid categories, only two are present so its still a single SVM)
# And select some of the more interesting features based on Box plots and scatter plots
diabetes_ipums <- clean_ipums %>% select(c("DIABETICEV", "AGE", "BMICALC", "ALCDAYSYR",
                                           "HOURSWRK", "SODAPNO", "FRIESPNO")) %>%
  na.omit();
write.csv(diabetes_ipums, file="output/preprocessing/data/selected_response_and_features.csv",
          row.names = FALSE);
save(diabetes_ipums, file="output/preprocessing/data/selected_response_and_features.RData");

print(head(diabetes_ipums));
```

```{r}

pca.result <- prcomp(diabetes_ipums[, -which(names(diabetes_ipums) == "DIABETICEV")],
                     center = TRUE, scale. = FALSE);
pca_df <- data.frame(
    comp1 = pca.result$x[,1], 
    comp2 = pca.result$x[,2], 
    DIABETICEV = diabetes_ipums$DIABETICEV)

plt <- ggplot(data = pca_df, mapping = aes(x = comp1, y = comp2, col=DIABETICEV,
                                           fill=DIABETICEV)) +
  geom_point() +
  labs(title="DIABETICEV in Principle Component Space",
       x="Principle Component 1",
       y="Principle Component 2",
       fill = "Diagnosed Diabetic",
       shape = "Diagnosed Diabetic",
       col = "Diagnosed Diabetic") +
  scale_fill_manual(values = c("1"="skyblue", "2"="coral", "3"="limegreen"),
                    labels = c("No", "Yes", "Borderline")) +
  scale_color_manual(values = c("1"="skyblue", "2"="coral", "3"="limegreen"),
                     labels = c("No", "Yes", "Borderline"));
print(plt);
ggsave(filename="output/eda/plots/response_class_in_pca_space.png", plot=plt, width = 8,
              height = 8, dpi="print");
```
